\chapter*{Acknowledgments}\normalsize
  \addcontentsline{toc}{chapter}{Acknowledgments}
%\pagestyle{plain}

\mquote{This work has been mainly supported by a scholarship of \textit{Deutsche Forschungsgemeinschaft (DFG)} as part of the \textit{Graduiertenkolleg "Systemintegration f\"{u}r Ubiquit\"{a}res Rechnen in der Informationstechnik" (GK 749)}, headed by Prof. Dr. Dr. h.c. mult. Manfred Glesner.}{}


This book has been based on a revised version of my Ph.D. thesis done at the Software Technology Group, at Darmstadt University of Technology and most of my acknowledgments go to the people that made the thesis possible. Doing a Ph.D. is a long project that one has to go through alone. However, no project of this size is possible without the help, valuable advice, and comments from a lot of other people. Especially my supervisor, Prof. Dr. Mira Mezini, provided continuous advice and reviewed parts of the material more than once. Her insistence that I care that the others understand what I say, has greatly contributed to the quality of this book. Finishing the dissertation would not have been possible without her continuous support.

The adventure of doing a Ph.D. started for me in 2000, when I went to the University of Siegen as part of a DAAD\footnote{Deutscher Akademischer Austauschdienst / German Academic Exchange Service, \url{http://www.daad.de}}  project, made possible by Prof. Dr.
Betim \c{C}i\c{c}o and Dr. Jochen M\"{u}nch, while I was waiting to obtain a working visa for the USA (thanks to Dr. Patrick Ball). The visa took longer than expected, and my professor in Siegen, Prof. Dr. Bernd Freisleben, \textit{somehow} convinced me to drop (in agreement with the employer - thanks to Marc Beneteau) the work possibility I had for the US, and start the research work toward a Ph.D. in Germany instead. I am grateful now to Prof. Dr. Freisleben for his support and for helping me to apply for a DFG\footnote{Deutschen Forschungsgemeinschaft / German Research Foundation, \url{http://www.dfg.de/}} scholarship at the Darmstadt University of Technology (TUD). Without his initial advice and support my Ph.D. would not exist.

I am also grateful to all my former colleagues at the Software Technology Group at the TUD, especially to Sven Kloppenburg, Dr. Michael Eichberg, Dr. Klaus Ostermann, Dr. Michael Haupt, Vaidas Gasiunas, Christoph Bockisch, and Thorsten Sch\"{a}fer, for their comments, reviews, and discussions. Eniana Mustafaraj and Elvis Papalilo from the University of Marburg reviewed some of the early drafts. Oliver Liemert implemented most of the prototype container framework for J2ME as part of his master diploma thesis. The anonymous reviewers of my conference and journal papers, upon which most parts of my thesis were based, provided many valuable comments and insights. I also want to thank Prof. Dr. Uwe A{\ss}mann for accepting to serve as co-reviewer of my thesis and for his comments on \kr{ch04}.

While it is difficult to write a technical document that everyone can understand, I hope most of the readers will enjoy the topics addressed in this book. Usually a Ph.D. is evaluated for the new contribution that it brings in a very narrow area of science. However, on the personal level the real value of a Ph.D. is the gained overall knowledge. I used the time and the university environment not only to contribute, but also to learn new things every day. The papers and the books I have read, including a very interesting one from the co-reviewer of my thesis, Prof. Dr. Uwe A{\ss}mann, the international conferences I have attended, the people I have met, the presentations I have seen, and the \textit{Graduiertenkolleg} I was part of, have enabled me to obtain a deeper understanding of the technologies discussed in this book. I am also thankful to Prof. Dr. Alejandro Buchmann for his guidance during the Graduiertenkolleg workshops. I would like to thank the additional members of the thesis Examination Committee, Prof. Dr. Thomas K\"{u}hne, and Prof. Dr. Andy Sch\"{u}rr for their reviews and comments. Many thanks go also to  Gudrun J\"{o}rs for helping me to organize a nice final buffet and for her administrative support during my time in Darmstadt.

Last but not least, I want to thank \textit{Ana}, for her unconditional support during the extremely stressful and energy draining years of my PhD. I am not sure whether I would have survived finishing my work without her. She has been simply living for me, and there are no words that I can find to describe my endless gratitude to her.

Finally, I want thank my publisher Bettina Wahlen from VDM Verlag Dr. M\"{u}ller e.K (VDM Publishing Hause) that made this book possible for a larger audience.

\vspace{0.5cm}
\noindent W\"{u}rzburg, 2007